---
title: "Prediction of SNP impact using convolutional neural networks."
author: "Raphael Mourad, Maître de Conférences, Université Paul Sabatier"
date: "10/02/2021"
output: html_document
---

## SETUP PROJECT FOLDER

Set up your project folder to your folder "atelier_INSERM".
```{r, setup}
knitr::opts_knit$set(root.dir = "/home/mourad/Downloads/Deep-Learning-for-Genomics-Training-main/")
knitr::opts_chunk$set(echo = TRUE)
```

## AIM OF THE SCRIPT

This script is used to predict the impact of a SNP on protein binding or chromatin accessibility for instance. For this purpose, trained convolutional neural networks are used.

## LOAD LIBRARIES

Load R librairies.
Remove suppressPackageStartupMessages() for debugging!
```{r, libraries}
suppressPackageStartupMessages(library(BSgenome.Hsapiens.UCSC.hg19))
suppressPackageStartupMessages(library(GenomicRanges))
suppressPackageStartupMessages(library(Biostrings))
suppressPackageStartupMessages(library(tensorflow))
suppressPackageStartupMessages(library(keras))
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(gplots))
```

## LOAD FUNCTIONS

```{r, function}
suppressPackageStartupMessages(source("scriptR/functions.R"))
```

## SOME PARAMETERS

Define parameters for the analysis:  
- peakSize is the resized peak size. For machine/deep learning, it is easier to work with peaks have the same size.  
- kpeaks is the number of best peaks used. Note that all peaks can also be used.  
- expe is the ChIP-seq experiment. For instance, CTCF for CTCF ChIP-seq or POL2 for RNA Polymerase 2 ChIP-seq experiment. 
```{r, parameters}
peakSize=201
kpeaks=4000
expe="CTCF"
DNAletters=c("A","C","G","T")
vocab_size=length(DNAletters)
```


## DEEP LEARNING PREDICTIONS OF SNP EFFECTS

rs907091 and rs9303277 were found in a GWAS of childhood acute lymphoblastic leukemia  
Article: https://www.nature.com/articles/s41467-017-02596-9 (supplementary file)  
rs907091 rs9303277 are also eQTLs in blood (https://www.gtexportal.org/home/snp/rs9303277)  
```{r, snp1}
SNP.GR1=GRanges("chr17",IRanges(c(37921742,37976469,38029120)),ref=c("C","C","C"),alt=c("T","T","G"),strand='+',
	rs=c("rs907091","rs9303277","rs12936231"),OR_GWAS=c(1.17,0.86,0.86), Effect_eQTL=c(0.32,-0.33,-0.34))
```

rs7329174 was found from GWAS catalog. It's associated with Crohn's disease and located in intron.
```{r, snp2}
SNP.GR2=GRanges("chr13",IRanges(c(41558110)),ref=c("A"),alt=c("G"), strand='+', rs=c(" rs7329174"),OR_GWAS=c(1.27))
```

rs142811167 and rs10821936 were found from GWAS catalog. It's associated with leukemia and located in intron.
```{r, snp3}
SNP.GR3=GRanges(c("chr11","chr10"),IRanges(c(7719585,63723577)),ref=c("C","T"),alt=c("T","C"), strand='+', rs=c("rs142811167","rs10821936"),OR_GWAS=c(87,1.8))
```

```{r, concat_snps}
SNP.GR=c(SNP.GR1,SNP.GR2,SNP.GR3)
print(SNP.GR)
```

Extract sequences surrounding the SNPs
```{r, getseq}
genomeVersion=BSgenome.Hsapiens.UCSC.hg19
region.GRr=resize(SNP.GR,peakSize,fix="center")
region.seq=as.character(getSeq(genomeVersion, names=seqnames(region.GRr),
     start=start(region.GRr), end=end(region.GRr)))
```


Make sequences with reference alleles
```{r, refseq}
SNPposrel=ceiling((peakSize+1)/2)
region.seqRef=region.seq
substring(region.seqRef,SNPposrel,SNPposrel)=as.character(region.GRr$ref)
names(region.seqRef)=paste0("seq",1:length(region.seqRef))
print(region.seqRef)
writeXStringSet(DNAStringSet(region.seqRef),filepath=paste0("data/fasta/sequences_SNPref_hg19.fa"))
```

Make sequences with alternative alleles
```{r, altseq}
region.seqAlt=region.seq
substring(region.seqAlt,SNPposrel,SNPposrel)=as.character(region.GRr$alt)
names(region.seqAlt)=paste0("seq",1:length(region.seqAlt))
print(region.seqAlt)
writeXStringSet(DNAStringSet(region.seqAlt),filepath=paste0("data/fasta/sequences_SNPalt_hg19.fa"))
```







